Chapter 16: Particle-laden gravity currents: the lock-release slumping regime at the laboratory scale

### 2024-03-03 paper/main.tex
Finishing up review and sending to Laurent

@@

It shall be noted that this dependency of the slumping velocity on the bottom slope exists despite the slumping equilibrium being unmodified by the presence of a slope-induced acceleration (see section~\ref{sec:intro} for details). As suggested in \citet{Gadal2023}, the latter acceleration may instead affect the complicated dynamics of the early transient acceleration phase, involving strong vertical motions and significant interfacial friction \citep{Cantero2007}.
%
One can write an energetic balance between the initial state and the end of this early acceleration phase \citep{Gadal2023}
\begin{equation}
	\label{eq:energetic_balance}
	\left[\frac{1}{2}\rho_{0}u_{\text c}^{2} - B \Delta\rho g \sin\alpha L\right] - A \Delta\rho g \cos\alpha h_{0} =  - \frac{1}{2}c_{\text d}\rho_{0}u_{\text c}^{2} \frac{L}{h_{0}},
\end{equation}
where $A$, $B$ and $c_{\text d}$ are scaling constants, and $L$ accounts for the alongslope distance over which the current moves during this phase. \citet{Cantero2007} found it to be independent of the lock aspect ratio or initial buoyancy, $L = 0.3 h_{0}$. Hence, we use here $L = D h_{0}$ with $D$ constant, which results in
\begin{equation}
	\label{eq:Fr_th}
	\mathcal{F}r = \frac{U_{\text c}}{U_0} = \frac{Fr_0}{\sqrt{1 + C_{\text D}}} \sqrt{1 + \frac{\tan\alpha}{S}},
\end{equation}
where $Fr_0$, $C_{\text D}$ and $S$ are new constants depending on $A$, $B$, $c_{\text d}$ and $D$. As shown in Fig.~\ref{fig:fig4}, this simple model is able to reproduce the observed variation of the dimensionless velocity in the bottom slope in the range [$5^\circ$, $35^\circ$].
%
For $\alpha < 5^\circ$, experimental data points are significantly scattered, and \eqref{eq:Fr_th} almost systematically overpredicts the $\mathcal{F}r$ values. A plausible explanation for this discrepancy could be the difference in the width-to-depth ratio of the flumes (and associated wall dissipation) and in aspect ratios of the lock, which are not addressed in the energetic balance leading to \eqref{eq:Fr_th}.
%
For $\alpha > 35^\circ$, the measured dimensionless velocities ($\mathcal{F}r$) slowly stand above equation (\ref{eq:Fr_th}) curve. In this region of large slope angles, the equilibrium between the forces likely becomes a lot more in favor of the so-called slope effect than the pressure gradient. Consequently, the scaling constants may have to be adapted to this major change in the influence of the forces.

\begin{figure*}[ht]
	\centering
	\includegraphics{figure4.pdf}
	\caption{\textbf{Influence of the bottom slope} on the current Froude number. Plain symbols correspond to $\phi < 0.45$, while transparent symbols are for $\phi \geq 0.45$ (see section~\ref{sec:influence_phi} for details on the influence of $\phi$). The black plain line indicates the prediction of \eqref{eq:Fr_th} for $Fr_{0} = 0.5$, $C_{\text D} = 0.4$ and $S = 0.25$. The black dashed line indicates the fit to the data from the numerical simulation of \citet{Birman2007} for saline homogeneous gravity currents.}
	\label{fig:fig4}
\end{figure*}

@@

\begin{figure*}[ht]
	\centering
	\includegraphics{figure5.pdf}
	\caption{\textbf{Influence of particle settling.} Current Froude number $\mathcal{F}r$ and attenuation $1/\tau$ as a function of the ratio between the Stokes number and the lock aspect-ratio, for different ranges of bottom slopes. Plain symbols correspond to $\phi < 0.45$, while transparent symbols are for $\phi \geq 0.45$ (see section~\ref{sec:influence_phi} for details on the influence of $\phi$). Dotted lines mark either the average of points with $\phi < 0.45$ (a, c, e, g) or the 0 baseline (b, d, f, h). On (e, f), the dashed line indicates the average value for homogeneous saline currents ($St \equiv 0$) of dataset 1. On (b, d, f, h), the black line is $1/\tau = 0.27 (St - 0.015)$.}
	\label{fig:fig5}
\end{figure*}

@@

\begin{figure}
	\centering
	\includegraphics{figure6.pdf}
	\caption{\textbf{Influence of particle volume fraction.} Current Froude number as a function of the initial particle volume fraction for (a) $\alpha \approx 0^\circ$, and (b) $\alpha \approx 45^\circ$. The inset in (a) shows the plot in linear scale. The black line indicates the prediction of \eqref{eq:Fr_th_eta} for $Fr_{0} = 0.5$, $C_{\text D} = 0.4$, $S = 0.25$ and $Re_{\text c} = 500$.}
	\label{fig:fig6}
\end{figure}

@@

The energetic balance presented in section~\ref{sec:influence_slope} can be adapted to include viscous dissipation effects, by replacing the constant $c_{\text d}$ in \eqref{eq:energetic_balance} by:
\begin{equation}
	c_{\text d}\left(1 + \frac{E}{\mathcal{R}e}\frac{U_0}{U_{\text c}}\frac{\eta_{\text eff}}{\eta_{\text f}} \right),
\end{equation}
where $E$ is a new scaling constant, and $\eta_{\text eff}$ is the effective suspension viscosity accounting for the presence of particles. This results in
\begin{equation}
	\label{eq:Fr_th_eta}
	\begin{split}
		\mathcal{F}r & =  \frac{1}{1 + C_{\text D}} \left[-\frac{Re_{\text c}}{\mathcal{R}e}\frac{\eta_{\text eff}}{\eta_{\text f}} \right. \\
		&\left. +\sqrt{\left(\frac{Re_{\text c}}{\mathcal{R}e}\frac{\eta_{\text eff}}{\eta_{\text f}} \right)^{2} + Fr_0^{2} \left(1 + \left[1 + C_{\text D}\right]\frac{\tan\alpha}{S} \right)}  \right],
	\end{split}
\end{equation}
where $Re_{\text c}$ is a new constant. The effect of the particles on the effective viscosity is based on the correlation proposed by \citet{krieger1959mechanism} and validated in many configurations:
\begin{equation}
	\label{eq:Krieger}
	\frac{\eta_{\text eff}}{\eta_{\text f}} = \left(1 - \frac{\phi}{\phi_{\text m}}\right)^{-(5/2)\phi_{\text m}},
\end{equation}
where $\phi_{\text m} = 0.585$ is the maximum volume fraction \citep{boyer2011unifying}.
%
As shown in figure~\ref{fig:fig6}(a), the combination of equations \eqref{eq:Fr_th_eta} and \eqref{eq:Krieger} can reproduce quantitatively the observed influence of $\phi$ on the dimensionless velocity for $\alpha=0^\circ$. This agreement does not depend on the choice of the viscosity model, and using other available rheological models from the literature \citep{stickel2005fluid, boyer2011unifying} leads to the same conclusions.

>>>>

\begin{figure*}
	\centering
	\includegraphics{figure4.pdf}
	\caption{\textbf{Influence of the bottom slope} on the current Froude number. Plain symbols correspond to $\phi < 0.45$, while transparent symbols are for $\phi \geq 0.45$ (see section~\ref{sec:influence_phi} for details on the influence of $\phi$). The black plain line indicates the prediction of \eqref{eq:Fr_th} for $Fr_{0} = 0.45$, $C_{\text D} = 0.02$ and $S = 0.25$. The black dashed line indicates the fit to the data from the numerical simulation of \citet{Birman2007} for saline homogeneous gravity currents.}
	\label{fig:fig4}
\end{figure*}

@@

It shall be noted that this dependency of the slumping velocity on the bottom slope exists despite the slumping equilibrium being unmodified by the presence of a slope-induced acceleration (see section~\ref{sec:intro} for details). As suggested in \citet{Gadal2023}, the latter acceleration may instead affect the complicated dynamics of the early transient acceleration phase, involving strong vertical motions and significant interfacial friction \citep{Cantero2007}.
%
One can write an energetic balance for the entire current between its initial state and the end of this early acceleration phase \citep{Gadal2023}
\begin{equation}
	\label{eq:energetic_balance}
	\left[\frac{1}{2}\rho_{0}u_{\text c}^{2} - B \Delta\rho g \sin\alpha L\right] - A \Delta\rho g \cos\alpha h_{0} =  - \frac{1}{2}c_{\text d}\rho_{0}u_{\text c}^{2} \frac{L}{h_{0}},
\end{equation}
where $A$ and $B$ are scaling constants, $c_{\text d}$ is a drag coefficient typically on the order of $10^{-2}$ -- $10^{-3}$ \citep{hogg2004effects}, and $L$ accounts for the alongslope distance over which the current moves during this phase. \citet{Cantero2007} found it to be independent of the lock aspect ratio or initial buoyancy, $L = 0.3 h_{0}$. Hence, we use here $L = D h_{0}$ with $D$ constant, which results in
\begin{equation}
	\label{eq:Fr_th}
	\mathcal{F}r = \frac{U_{\text c}}{U_0} = \frac{Fr_0}{\sqrt{1 + C_{\text D}}} \sqrt{1 + \frac{\tan\alpha}{S}},
\end{equation}
where $Fr_0$, $C_{\text D}$ and $S$ are new constants depending on $A$, $B$, $c_{\text d}$ and $D$. Note that $C_{\rm d}$ has little influence if $c_{\text d}$ is kept within the reasonable values mentioned previously, and $Fr_0$ is strongly restricted by the values in $\theta = 0^\circ$, such that the only truly adjustable parameter is $S$. As shown in Fig.~\ref{fig:fig4}, this simple dimensional model is able to reproduce the observed variation of the dimensionless velocity in the bottom slope in the range [$5^\circ$, $35^\circ$] while keeping the scaling constants within reasonable values.
%
For $\alpha < 5^\circ$, experimental data points are significantly scattered, and \eqref{eq:Fr_th} almost systematically overpredicts the $\mathcal{F}r$ values. A plausible explanation for this discrepancy could be the difference in the width-to-depth ratio of the flumes (and associated wall dissipation) and in aspect ratios of the lock, which are not addressed in the energetic balance leading to \eqref{eq:Fr_th}.
%
For $\alpha > 35^\circ$, the measured dimensionless velocities ($\mathcal{F}r$) slowly stand above equation (\ref{eq:Fr_th}) curve. In this region of large slope angles, the equilibrium between the forces likely becomes a lot more in favor of the so-called slope effect than the pressure gradient. Consequently, the scaling constants may have to be adapted to this major change in the influence of the forces.

@@

\begin{figure*}
	\centering
	\includegraphics{figure5.pdf}
	\caption{\textbf{Influence of particle settling.} Current Froude number $\mathcal{F}r$ and attenuation $1/\tau$ as a function of the ratio between the Stokes number and the lock aspect-ratio, for different ranges of bottom slopes. Plain symbols correspond to $\phi < 0.45$, while transparent symbols are for $\phi \geq 0.45$ (see section~\ref{sec:influence_phi} for details on the influence of $\phi$). Dotted lines mark either the average of points with $\phi < 0.45$ (a, c, e, g) or the 0 baseline (b, d, f, h). On (e, f), the dashed line indicates the average value for homogeneous saline currents ($St \equiv 0$) of dataset 1. On (b, d, f, h), the black line is $1/\tau = 0.27 (St - 0.015)$.}
	\label{fig:fig5}
\end{figure*}

@@

\begin{figure}
	\centering
	\includegraphics{figure6.pdf}
	\caption{\textbf{Influence of particle volume fraction.} Current Froude number as a function of the initial particle volume fraction for (a) $\alpha \approx 0^\circ$, and (b) $\alpha \approx 45^\circ$. The inset in (a) shows the plot in linear scale. The black line indicates the prediction of \eqref{eq:Fr_th_eta} for $Fr_{0} = 0.45$, $C_{\text D} = 0.02$, $S = 0.25$ and $Re_{\text c} = 500$.}
	\label{fig:fig6}
\end{figure}

@@

The energetic balance presented in section~\ref{sec:influence_slope} can be adapted to include viscous dissipation effects, by replacing the constant $c_{\text d}$ in \eqref{eq:energetic_balance} by:
\begin{equation}
	c_{\text d}\left(1 + \frac{E}{\mathcal{R}e}\frac{U_0}{U_{\text c}}\frac{\eta_{\text eff}}{\eta_{\text f}} \right),
\end{equation}
where $E$ is a new scaling constant, and $\eta_{\text eff}$ is the effective suspension viscosity accounting for the presence of particles. This results in
\begin{equation}
	\label{eq:Fr_th_eta}
	\begin{split}
		\mathcal{F}r & =  \frac{1}{1 + C_{\text D}} \left[-\frac{Re_{\text c}}{\mathcal{R}e}\frac{\eta_{\text eff}}{\eta_{\text f}} \right. \\
		&\left. +\sqrt{\left(\frac{Re_{\text c}}{\mathcal{R}e}\frac{\eta_{\text eff}}{\eta_{\text f}} \right)^{2} + Fr_0^{2} \left(1 + \left[1 + C_{\text D}\right]\frac{\tan\alpha}{S} \right)}  \right],
	\end{split}
\end{equation}
where $Re_{\text c}$ is a new constant. The effect of the particles on the effective viscosity is based on the correlation proposed by \citet{krieger1959mechanism} and validated in many configurations:
\begin{equation}
	\label{eq:Krieger}
	\frac{\eta_{\text eff}}{\eta_{\text f}} = \left(1 - \frac{\phi}{\phi_{\text m}}\right)^{-(5/2)\phi_{\text m}},
\end{equation}
where $\phi_{\text m} = 0.585$ is the maximum volume fraction \citep{boyer2011unifying}.
%
Here, the only adjustable parameter is $Re_{\text c}$ as all the other parameters have been fixed previously. As shown in figure~\ref{fig:fig6}(a), the combination of equations \eqref{eq:Fr_th_eta} and \eqref{eq:Krieger} can reproduce quantitatively the observed influence of $\phi$ on the dimensionless velocity for $\alpha=0^\circ$. This agreement does not depend on the choice of the viscosity model, and using other available rheological models from the literature \citep{stickel2005fluid, boyer2011unifying} leads to the same conclusions.

### 2024-03-03 paper/reply_to_referees/first_review/reply.tex
Finishing up review and sending to Laurent

\noindent We thank you very much for considering our chapter ``Chapter 16: Particle-laden gravity currents: the lock-release slumping regime at the laboratory scale", by C. Gadal \emph{et al}, for publication in a book for the \emph{Geophysical Monograph Series}. We are grateful for the thoughtful referee reports, and for their comments and suggestions. \\

@@

\begin{itemize}
    \item \MinorReply{Introduction}{The literature, although rich, lacks a number of contributions of some importance, mainly because they are themselves experimental. See, for example, the work of Zemach et al. 2017, which deals with some more complex configurations, with non-rectangular geometry, with stratified fluid, possibly in the partial depth lock release configuration. The various contributions should be catalogued and described in an appropriate way.}{The introduction has been rewritten according to various comments of the referees and other readers. Some references have been added. Note that we can not provide an extensive catalog of the complete literature related gravity currents in such an introduction, bound to have a reasonable length. \tocheck{TO DO.}}

    \item \MinorReply{Figure 1}{the q-v panels look out of place and the sequence is horizontal while the experiments are vertical. It is recommended that they be placed in a separate figure.}{\tocheck{There are already many figures compared to the text length. As this is properly described in the caption, we prefer to leave it as one figure presenting all configurations.}}

    \item \MinorReply{Figure 3a}{Figure 3a is not meaningful due to the large amount of data shown. It is advisable to scale it down in dimensionless form, as is done in Figure 3b for 'selected runs', which should be specified.}{\tocheck{Do we do it?}}

    \item \MinorReply{Figure 3b}{the alpha symbol is flanked by an arrow which appears to be meaningless. I assume that the arrow should be reversed.}{\tocheck{The upward arrow flanking the alpha symbol means that alpha is increasing. We prefer to leave it this way.}}

@@

\begin{itemize}
    \item \MinorReply{Line 64}{Line 64 is unclear}{Done.}
    \item \MinorReply{Line 78}{The sentence starting in line 78 is too convoluted. Needs restructuring.}{Done.}
    \item \MinorReply{Line 102}{what does it mean by $h_c$ can have different scaling?}{The two scalings are explained in the two following sentences.}
    \item \MinorReply{Lines 104--107}{It is unclear what the authors mean gt detachment. Also “similarly” may need to change to “similar”}{\laurent{laurent}. Je serai pour enlever le scaling des colonnes lourdes non-Boussinesq. C'est pas clair pour moi, et je pense que ce que l'on dit est faux : dans un dam-break de courant de gravité Bussinesq, on a le courant qui reste attaché à la colonne au début tout en ayant $h_{\rm c} \propto H_{0}$.}
    \item \MinorReply{Section 1.1}{gives several different expressions for $X_t$. Some clarification as to where each apply will be useful.}{The physical process associated to each of the expression of $X_{t}$ is clearly stated each time. Hence, each expression applies when this process dominates the other. Characterizing these transitions is exactly one of the purposes of the data presented here. This is now specified in the text. \tocheck{TO DO}.}
    \item \MinorReply{Lines 133-134}{where do length and time scale of particles play a role?}{}
    \item \MinorReply{Line 211-212}{what does it mean 2D vertical and turbulence averaged? Are these 2D RANS simulations? Are the 2 coordinates horizontal, since vertical is averaged?}{\julien{julien}}
    \item \MinorReply{In figure 2}{the color of the symbols outline must be the same. Otherwise it is hard to relate to the color legend}{The color of the outline of the symbols is used to show the difference between two different kinds of runs, as explained in the caption. We prefer to leave it this way.}
    \item \MinorReply{figure 2}{some of the volume fraction data goes to 1.}{\marie{marie}}
    \item \MinorReply{}{Curve fits with such a large data scatter, requires some justification}{}
    \item \MinorReply{Line 461}{even even}{Done.}
    \item \MinorReply{Line 466}{This should be an important….}{Done.}
\end{itemize}

>>>>

\noindent We thank you very much for considering our chapter ``Chapter 16: Particle-laden gravity currents: the lock-release slumping regime at the laboratory scale", by C. Gadal \emph{et al}, for publication in a book for the \emph{Geophysical Monograph Series}. We are grateful for the thoughtful referee reports and for their comments and suggestions. \\

@@

\begin{itemize}
    \item \MinorReply{Introduction}{The literature, although rich, lacks a number of contributions of some importance, mainly because they are themselves experimental. See, for example, the work of Zemach et al. 2017, which deals with some more complex configurations, with non-rectangular geometry, with stratified fluid, possibly in the partial depth lock release configuration. The various contributions should be catalogued and described in an appropriate way.}{The introduction has been rewritten according to various comments of the referees and other readers. Some references have been added. Note that we can not provide an extensive catalog of the complete literature related to gravity currents in such an introduction, bound to have a manuscript of reasonable length. We have then focused on the literature that shares the most scientific objectives with the present contribution.}

    \item \MinorReply{Figure 1}{the q-v panels look out of place and the sequence is horizontal while the experiments are vertical. It is recommended that they be placed in a separate figure.}{There are already many figures compared to the text length, and we do not wish to emphasize the numerical simulations by adding a dedicated figure. Hence, we modified figure 1 to better separate the q-v panels from the rest.}

    \item \MinorReply{Figure 3a}{Figure 3a is not meaningful due to the large amount of data shown. It is advisable to scale it down in dimensionless form, as is done in Figure 3b for 'selected runs', which should be specified.}{Done.}

    \item \MinorReply{Figure 3b}{the alpha symbol is flanked by an arrow which appears to be meaningless. I assume that the arrow should be reversed.}{The flanking arrows have been removed, and the figure caption has been modified accordingly.}

@@

\begin{itemize}
    \item \MinorReply{Line 64}{Line 64 is unclear}{Done.}

    \item \MinorReply{Line 78}{The sentence starting in line 78 is too convoluted. Needs restructuring.}{Done.}

    \item \MinorReply{Line 102}{what does it mean by $h_c$ can have different scaling?}{This sentence has been removed, and this part of the text re-written for the sake of clarity.}

    \item \MinorReply{Lines 104--107}{It is unclear what the authors mean gt detachment. Also “similarly” may need to change to “similar”}{This part of the text has been re-written for the sake of clarity.}

    \item \MinorReply{Section 1.1}{gives several different expressions for $X_t$. Some clarification as to where each apply will be useful.}{The physical process associated with each expression of $X_{t}$ is stated along the section when $X_{t}$ is defined. Now, the first transition to occur will depend on the prevailing mechanism, and as such on where the experimental run is in the parameter space. This is now clarified in the introduction.}

    \item \MinorReply{Lines 133-134}{where do length and time scale of particles play a role?}{This sentence has been modified by adding examples.}

    \item \MinorReply{Line 211-212}{what does it mean 2D vertical and turbulence averaged? Are these 2D RANS simulations? Are the 2 coordinates horizontal, since vertical is averaged?}{\julien{julien}}

    \item \MinorReply{In figure 2}{the color of the symbols outline must be the same. Otherwise it is hard to relate to the color legend}{The color of the outline of the symbols is used to show the difference between two different kinds of runs, as explained in the caption. We prefer to leave it this way.}

    \item \MinorReply{figure 2}{some of the volume fraction data goes to 1.}{\marie{marie}}

    \item \MinorReply{}{Curve fits with such a large data scatter, requires some justification}{Here, we do not fit a proper analytical model or check the validity of its prediction against the data. We merely aim to show that a simple dimensional model incorporating the required mechanisms is able to reproduce the observed variation of the data while keeping the scaling constants within reasonable values. In addition, we now clarify that the only true adjustable parameters are $S$ for the slope effect and $Re_{c}$ for the effect of the volume fraction, as $C_{\rm d}$ and $Fr_{0}$ are strongly bound by values in the literature.}

    \item \MinorReply{Line 461}{even even}{Done.}

    \item \MinorReply{Line 466}{This should be an important….}{Done.}
\end{itemize}
